Our Galactic Midplane Is Probably Wrong, and Not by a Little

### 2018-12-14 paper/ms.tex
Update paper

\shortauthors{Beane et al.}

@@

\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\z}{z}

@@

\title{Our Galactocentric Coordinate System Is Probably Wrong, and Not by a
Little}

@@

Since the Milky Way is not truly axisymmetric, it is not clear the best way to
define the galactocentric coordinate system. In this work, we consider one
possible physically motivated approach to defining the galactocentric
coordinate system, and show that it can differ from conventional methods of
frame measurements.

@@

@@

\subsection{Local Midplane} \label{sec:local_midplane}
We now turn to defining the local midplane that an observer might measure if
they were situated in each of these galaxies. We place our imaginary observer
at the solar radius of $8.2\,\kpc$ and azimuth $\phi$, and consider stars
within a cylinder perpendicular to the fiducial disk

@@

To allow for potential small inaccuracies in the fiducial coordinate systems,
we also subtract the best fit $\text{A} \sin{\phi + \text{B}} + C$ from the
midplane as a function of azimuth. For $\text{A}$, the values are $-165$,
$45$, $9\,\pc$ and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$
for m12i, m12f, and m12m, respectively. For the assumed solar radius of
$8.2\,\kpc$, we can approximate the angle offset for the $z$-axis from the
values of $\text{A}$ --- we compute $1.15$, $0.31$, and $0.06\,\deg$ for m12i,
m12f, and m12m. These angle offsets are consistent with the values given in
\citet{2018arXiv180610564S} for the difference between the $z$-axis as defined
by the gas and stars.

@@

\subsection{Dynamical Midplane}
We now seek to define the midplane in a dynamically consistent manner, which
we refer to as the ``dynamical midplane''. By this, we mean we would like to
define the midplane in a manner consistent with some sort of dynamical
constraint. Our method, which we describe now, is just one of many possible
ways the dynamical midplane can be defined.

We define the dynamical midplane based on the principle that perfectly
circular, planar orbits will trace the midplane. Once the circular, planar
orbits have been identified, the procedure for finding the dynamical midplane
is relatively simple. From the measurement of each star's orbit, the midplane
of the galaxy can be computed at $z=0$ easily. Furthermore, we will find that
each orbit also gives a measure of the: dynamical center of the galaxy,
velocity of the galaxy, and acceleration of the galaxy.

We pause briefly to consider general difficulties in measuring stellar orbits
in a cosmological simulation. It may seem reasonable to place the galaxy in an
inertial reference frame and then to simply measure the positions of each star
particle over time. However, it is not generally the case that the center of
mass of the stellar disk experiences a vanishing acceleration. Indeed, in a
re-run of the simulations presented here\footnote{The re-run contained a
cosmic ray bug fix which is unimportant here, but the re-run also output the
acceleration of each particle, which is not the case for the simulations in
this work.}, we measure a total acceleration of $\sim 0.4\,\kms/\text{Myr}$
for stars with cylindrical radii within $0.5\,\kpc$ of $8.2\,\kpc$ and
vertical heights less than $0.4\,\kpc$. Assuming a constant acceleration, over
$22\,\Myr$ --- the time spacing of snapshots considered in this work --- this
results in a $\sim90\,\pc$ shift of the disk. Since the offset scales with
$t^2$, this presents a significant problem for measuring stellar orbits over
longer timescales.

Another option is to recenter and reorient the galaxy at each snapshot.
However, the centering and orientation procedures are not without
stochasticity themselves. This stochasticity can easily dominate the magnitude
of the orbit. For instance, a shift in the direction of the $z$-axis by a mere
$0.7\degree$ results in a shift in the midplane of $100\,\pc$ at a radius of
$8.2\,\kpc$. We attempted this procedure in various ways (i.e. exploring
different centering and orientation procedures), but were not able to cleanly
measure the orbits of disk stars.

Here we choose to simplify the problem by considering only perfectly circular,
planar orbits. Since these orbits have no phase-dependent structure, we can
work on shorter timescales, i.e. for orbits with significant radial or
vertical action a large fraction of their orbit would need to be sampled in
order to measure the orbit accurately. Because we know the analytic form of
these orbits, we can forward model the trajectory of each star particle in the
cosmological simulation. Working in comoving coordinates, we can write down
the trajectory of a star particle as

\beq
\bm{r}(t) = R(\theta, \phi, \psi) \left( (A\cos{\omega t}, B\sin{\omega t}, C t) + \bm{x_0} \right)
\eeq

\section{Coordinate System Gone Wrong} \label{sec:ref_frame}
We show and quantify the time evolution of measured actions that comes about
from inaccuracies in the galactocentric coordinate system.

\subsection{Cartoon} \label{ssec:cartoon}
We begin with a cartoon in Figure~\ref{fig:cartoon} to show that an inaccurate
determination of The solid gray line indicates a hypothetical ``true'' orbit.
The y-axis corresponds to the vertical height of the orbit and the x-axis to
either time or orbital phase. The dashed gray line is an incorrect midplane
which an observer uses to compute orbits.

\subsection{Quantification} \label{ssec:quant}
We now turn to classical orbital integration in a commonly assumed Milky Way
potential to quantify the time variation of computed actions for different
coordinate system errors.

We use the code \texttt{gala} v0.3 to perform orbit integrations and
conversion to action space \citep{2017JOSS....2..388P,Price-Whelan:2018},
which uses an action estimator technique presented by
\citet{2014MNRAS.441.3284S}. We use the default \texttt{MWPotential} as our
potential, which is based on the Milky Way potential available in
\texttt{galpy} \citep{2015ApJS..216...29B}. This potential includes a
Hernquist bulge and nucleus \citep{1990ApJ...356..359H}, a Miyamoto-Nagai disk
\citep{1975PASJ...27..533M}, and an NFW halo \citep{1997ApJ...490..493N}, and
is fit to empirically match some observations. We use the Dormand-Prince
8(5,3) integration scheme \citep{Dormand80:integrator}. We use a timestep of
$1\,\Myr$ and integrate for $5\,\Gyr$, corresponding to $\sim 20$ orbits for
a Sun-like star.

Using the observations of the orbit that this erroneous observer makes every
$\Myr$, we perform the same orbit integration procedure as before to compute
the actions our observer would make in the incorrect coordinate system over the
duration of the orbit. The erroneous computed actions are shown for the first
$\Gyr$ of the orbit in Figure~\ref{fig:one_orbit_wrong_ref}. Occasionally the
numerical scheme fails and very large actions are reported by \texttt{gala}
--- we perform a $5\sigma$ clip on each action to exclude such orbits, but
this only excludes a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.

We now

>>>>

\shortauthors{Beane et al.}
\shorttitle{Where the Hell Are We?}

@@

\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\z}{z}
\newcommand{\uth}{\textsuperscript{th}} % why is this so long in the first place??

@@

\title{Our Galactic Midplane Is Probably Wrong, and Not by a Little}

@@

Since the Milky Way is not truly axisymmetric, it is not clear the best way to
define the Galactocentric coordinate system. In this work, we focus on the midplane and, equivalently, the solar height.

@@

\section{Impact on Actions} \label{sec:ref_frame}
We first demonstrate the importance of measuring the Galactic midplane to very
high accuracy. We explore the consequence of an incorrect midplane in the
context of computing orbital actions of disk stars. An inaccuracy in the
midplane is only one potential systematic error in the way actions are
presently computed. The assumed axisymmetric model of the Galaxy may be a poor
description of the Galactic potential, and the parameters in this axisymmetric
model may be estimated incorrectly.

In this work, we assume both that an axisymmetric model of the Galaxy is a good description of the potential and that we have correctly determined the parameters of that model. We are only concerned with an inaccuracy

\subsection{Summary of Action Computation} \label{ssec:action_comp}
We begin by summarizing one way in which actions are computed in the
literature \Gus{cite}. One first assumes an axisymmetric model of the Galaxy.
This model may itself be a poor description of the Galactic potential,

We use the code \texttt{gala} v0.3 to perform orbit integrations and
conversion to action space \citep{2017JOSS....2..388P,Price-Whelan:2018},
which uses an action estimator technique presented by
\citet{2014MNRAS.441.3284S}. Another common alternative is \texttt{galpy}
\citep{2015ApJS..216...29B} using either direct orbit integration or the
\Gus{Stackel Fudge} approximation. We use the default \texttt{MWPotential} as
our potential, which is based on the Milky Way potential available in
\texttt{galpy} \citep{2015ApJS..216...29B}. This potential includes a
Hernquist bulge and nucleus \citep{1990ApJ...356..359H}, a Miyamoto-Nagai disk
\citep{1975PASJ...27..533M}, and an NFW halo \citep{1997ApJ...490..493N}, and
is fit to empirically match some observations. We use the Dormand-Prince
8(5,3) integration scheme \citep{Dormand80:integrator}. We use a timestep of
$1\,\Myr$ and integrate for $5\,\Gyr$, corresponding to $\sim 20$ orbits for a
Sun-like star.

\subsection{Cartoon} \label{ssec:cartoon}
We present a cartoon in Figure~\ref{fig:cartoon} to show that an inaccurate
determination of the midplane can lead to a phase-dependence of the {\em
observed} actions. The solid gray line indicates a hypothetical ``true''
orbit. The y-axis corresponds to the vertical height of the orbit and the
x-axis orbital phase. The dashed gray line is an incorrect midplane which an
observer uses to compute orbits.

\subsection{Quantification} \label{ssec:quant}
We now turn to classical orbital integration in a commonly assumed Milky Way
potential to quantify the time variation of computed actions for different
coordinate system errors. Our action computation procedure is described in
Section~\ref{ssec:action_comp}.

Using the observations of the orbit that this erroneous observer makes every
$\Myr$\footnote{In addition to being erroneous, we also assume our fictitious
observer is immortal.}, we perform the same orbit integration procedure as
before to compute the actions our observer would make in the incorrect
coordinate system over the duration of the orbit. The erroneous computed
actions are shown for the first $\Gyr$ of the orbit in
Figure~\ref{fig:one_orbit_wrong_ref}. Occasionally the numerical scheme fails
and very large actions are reported by \texttt{gala}
--- we perform a $5\sigma$ clip on each action to exclude such orbits, but
this only excludes a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.

We now repeat the procedure but continuously changing the error in the assumed
midplane. In Figure~\ref{fig:many_orbit_wrong_ref}, we report the 95\uth minus
5\uth

\begin{figure*}
% \plotone{fig/something.pdf}
\caption{Caption.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

\subsection{Local Midplane} \label{sec:local_midplane}
We now turn to defining the local midplane that an observer might measure if
they were situated in each of these galaxies. We place our imaginary observer
at the solar radius of $8.2\,\kpc$ and azimuth $\phi$, and consider stars
within a cylinder perpendicular to the fiducial disk.

@@

To allow for potential small inaccuracies in the fiducial coordinate systems,
we also subtract the best fit $\text{A} \sin{\phi + \text{B}} + C$ curve from
the midplane as a function of azimuth. For $\text{A}$, the values are $-165$,
$45$, $9\,\pc$ and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$
for m12i, m12f, and m12m, respectively. For the assumed solar radius of
$8.2\,\kpc$, we can approximate the angle offset for the $z$-axis from the
values of $\text{A}$ --- we compute $1.15$, $0.31$, and $0.06\,\deg$ for m12i,
m12f, and m12m. These angle offsets are consistent with the values given in
\citet{2018arXiv180610564S} for the difference between the $z$-axis as defined
by the gas and stars.

@@

% \subsection{Dynamical Midplane}
% We now seek to define the midplane in a dynamically consistent manner, which
% we refer to as the ``dynamical midplane''. By this, we mean we would like to
% define the midplane in a manner consistent with some sort of dynamical
% constraint. Our method, which we describe now, is just one of many possible
% ways the dynamical midplane can be defined.

% We define the dynamical midplane based on the principle that perfectly
% circular, planar orbits will trace the midplane. Once the circular, planar
% orbits have been identified, the procedure for finding the dynamical midplane
% is relatively simple. From the measurement of each star's orbit, the midplane
% of the galaxy can be computed at $z=0$ easily. Furthermore, we will find that
% each orbit also gives a measure of the: dynamical center of the galaxy,
% velocity of the galaxy, and acceleration of the galaxy.

% We pause briefly to consider general difficulties in measuring stellar orbits
% in a cosmological simulation. It may seem reasonable to place the galaxy in an
% inertial reference frame and then to simply measure the positions of each star
% particle over time. However, it is not generally the case that the center of
% mass of the stellar disk experiences a vanishing acceleration. Indeed, in a
% re-run of the simulations presented here\footnote{The re-run contained a
% cosmic ray bug fix which is unimportant here, but the re-run also output the
% acceleration of each particle, which is not the case for the simulations in
% this work.}, we measure a total acceleration of $\sim 0.4\,\kms/\text{Myr}$
% for stars with cylindrical radii within $0.5\,\kpc$ of $8.2\,\kpc$ and
% vertical heights less than $0.4\,\kpc$. Assuming a constant acceleration, over
% $22\,\Myr$ --- the time spacing of snapshots considered in this work --- this
% results in a $\sim90\,\pc$ shift of the disk. Since the offset scales with
% $t^2$, this presents a significant problem for measuring stellar orbits over
% longer timescales.

% Another option is to recenter and reorient the galaxy at each snapshot.
% However, the centering and orientation procedures are not without
% stochasticity themselves. This stochasticity can easily dominate the magnitude
% of the orbit. For instance, a shift in the direction of the $z$-axis by a mere
% $0.7\degree$ results in a shift in the midplane of $100\,\pc$ at a radius of
% $8.2\,\kpc$. We attempted this procedure in various ways (i.e. exploring
% different centering and orientation procedures), but were not able to cleanly
% measure the orbits of disk stars.

% Here we choose to simplify the problem by considering only perfectly circular,
% planar orbits. Since these orbits have no phase-dependent structure, we can
% work on shorter timescales, i.e. for orbits with significant radial or
% vertical action a large fraction of their orbit would need to be sampled in
% order to measure the orbit accurately. Because we know the analytic form of
% these orbits, we can forward model the trajectory of each star particle in the
% cosmological simulation. Working in comoving coordinates, we can write down
% the trajectory of a star particle as
% \beq\label{eq:helix}
% \bm{r}(t) = R(\theta, \phi, \psi) \left[ (A\cos{\omega t}, B\sin{\omega t}, C
% t) + \bm{x_0} \right]\text{,}
% \eeq
% where $R(\theta, \phi, \psi)$ is an Euler angle transformation, $A$, $B$, $C$,
% and $\omega$ are arbitrary constants, and $\bm{x_0}$ is an arbitrary 3-vector.
% This equation is a helix with an arbitrary constant and rotation. It describes
% the

% We minimize the summed squared difference between Equation~\ref{eq:helix} and
% the true trajectory of each star with $R$ within $1\,\kpc$ of $R_{\odot} =
% 8.2\,\kpc$ in each simulation. We consider minimum rms error (i.e. the typical
% error in the Equation~\ref{eq:helix} model in describing the true trajectory).
% This rms error characterizes how well the helix model we consider, which would
% be exactly correct if the overall motion of the galaxy is well-described by
% constant speed motion and the orbit of a particular store is exactly circular
% and planar.

% We plot a histogram of the rms error for each galaxy in
% Figure~\ref{fig:something}. As we estimated before, the total effect of the
% galaxy's acceleration is roughly $90\,\pc$ (\Gus{Check for m12f, m12m}). We
% find that this is small compared to the typical error, even for the stars
% whose trajectories are best modeled by Equation~\ref{eq:helix}.

% There are many reasons for the failure of this model. First, there could be no
% stars on circular orbits. This seems unlikely considering that typical
% distribution functions peak on perfectly circular, planar orbits
% \citep[e.g.][]{2013MNRAS.434..652T}. Second, the trajectory of the galaxy may
% not be well-described by constant-speed motion. This also seems unlikely,
% considering that the bulk acceleration of the stellar disk only results in a
% $\sim90\,\pc$ (\Gus{m12f, m12m}) deviation from constant-speed motion over the
% $22\,\Myr$ time we considered.

% Furthermore, in Figure~\ref{fig:something} we plot the center of mass of the
% galaxy (in comoving coordinates) for the $22\,\Myr$ timeframe described. The
% center of mass at each snapshot is computed using the iterative shrinking
% spheres method described in Section~\ref{ssec:cosmozoom}. The rms error on the
% constant-speed motion is $\sim\text{something}$, which we interpret as
% \Gus{something}.

% One potential explanation is that {\em no star can remain on a circular,
% planar orbit}. The order unity density variations may be strong enough such
% that each star's orbit is perturbed. If there are truly stars in the
% simulation on perfectly circular, planar orbits, then the magnitude of their
% acceleration over the last $22\,\Myr$ must be constant. Furthermore, their
% trajectory in acceleration space must be well-described by a rotation. In the re-runs,
